Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-06-03 manu_draft.Rmd
added seastrnd data and table

data(locs)
data(modprf)

@@

Table \@ref(tab:modprftab)

@@

```{r modsumtab}
tab <- tibble(
  GAM = c('`gam0`', '`gam1`', '`gam2`', '`gam6`'),
  Summary = c(
    'Linear trend with seasonality', 
    'Nonlinear trends seasonality (constrained knots)', 
    'Nonlinear trends wiht seasonality, plus interaction',
    'Nonelinear trends with seasonality (unconstrained knots)'
  ),
  Details = c(
    'A linear trend in time over the course of the time series (cyear); and Within-year seasonal fluctuations that follow the same pattern every year - e.g. large spring bloom and small fall bloom (`s(doy, bs = "cc")`)',
    'Treats cyear and doy same as gam0; and A smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
    'Treats cyear and doy same as gam1; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to gam6 (`ti(year, doy, bs = c("tp", "cc"))`)', 
    'Variant of gam1 with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
  )

@@

```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).  Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.  Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized. Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding.  Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
totab <- modprf %>% 
  group_by(station) %>% 
  mutate(
    AIC = AIC - max(AIC),
    minaic = AIC == min(AIC), 
    mingcv = GCV == min(GCV),
    maxr2 = R2 == max(R2)
  ) %>% 
  ungroup

@@

>>>>

data(locs)
data(modprf)
data(seastrnd)

source('R/funcs.R')

@@

Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab)

@@

```{r modsumtab}
tab <- tibble(
  GAM = c('`gam0`', '`gam1`', '`gam2`', '`gam6`'),
  Summary = c(
    'Linear trend with seasonality', 
    'Nonlinear trends seasonality (constrained knots)', 
    'Nonlinear trends wiht seasonality, plus interaction',
    'Nonlinear trends with seasonality (unconstrained knots)'
  ),
  Details = c(
    'A linear trend in time over the course of the time series (year); and Within-year seasonal fluctuations that follow the same pattern every year, e.g., large spring bloom and small fall bloom (`s(doy, bs = "cc")`)',
    'Treats year and doy same as `gam0`; and A smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
    'Treats cyear and doy same as `gam1`; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to `gam6` (`ti(year, doy, bs = c("tp", "cc"))`)', 
    'Variant of `gam1` with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
  )

@@

```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).  Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.  Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized. Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
totab <- modprf %>% 
  group_by(station) %>% 
  mutate(
    AIC = AIC - max(AIC),
    minaic = AIC == min(AIC), 
    mingcv = GCV == min(GCV),
    maxr2 = R2 == max(R2)
  ) %>% 
  ungroup

@@

```{r seastrndtab, tab.cap = 'Seasonal trends from `gam6` results by station and decade.  Results show the annual slope estimate as chlorophyll change per year for each seasonal period across years (JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September, OND: October, November, December).  Significant trends based on mixed-meta regression analysis are indicated with asterisks at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
totab <- seastrnd %>% 
  mutate(
    pval = p_ast(pval),
    pval = gsub('ns', '', pval), 
    yrcoef = round(yrcoef, 2)
    ) %>% 
  unite('yrcoef', yrcoef, pval, sep = '') %>% 
  spread(seas, yrcoef) %>% 
  mutate(
    station = ifelse(duplicated(station), '', station)
  )

boldjfm <- grep('\\*', totab$JFM)
boldamj <- grep('\\*', totab$AMJ)
boldjas <- grep('\\*', totab$JAS)
boldond <- grep('\\*', totab$OND)

tab <- flextable(totab) %>% 
  set_header_labels(
    station = 'Station', 
    yrs = 'Years'
  ) %>% 
  add_header_row(values = c('', '', 'Chlorophyll change (ug/l) per year', '', '', '')) %>% 
  merge_at(i = 1, j = 3:6, part = 'header') %>% 
  bold(i = boldjfm, j = 'JFM') %>% 
  bold(i = boldamj, j = 'AMJ') %>% 
  bold(i = boldjas, j = 'JAS') %>% 
  bold(i = boldond, j = 'OND') %>% 
  border_remove() %>% 
  hline_top(border = fp_border(color = 'black')) %>% 
  font(fontname = 'Times', part = 'all') %>%
  fontsize(size = 12, part = 'all') %>%   
  width(j = c(2), width = 1) 
tab
```
